Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2021-01-18 journal paper/content/related_work.tex
Minor space-saving.

\xhdr{Exploration.} Multi-armed bandits (\emph{MAB}) is an elegant and tractable abstraction for tradeoff between \emph{exploration} and \emph{exploitation}: essentially, between acquisition and usage of information. MAB problems have been studied for many decades by researchers from computer science, operations research, statistics and economics, generating a vast and multi-threaded literature.  The most relevant thread concerns the basic model of regret-minimizating bandits with stochastic rewards and no auxiliary structure (which is the problem faced by each principal in our model), see Appendix~\ref{app:bg} for background. This basic model has been extended in many different directions, with a considerable amount of work on each: \eg payoffs with a specific structure (\eg combinatorial, linear, convex or Lipschitz), payoff distributions that change over time, and auxiliary payoff-relevant signals.%
\footnote{There is a superficial similarity, in name only, between this paper and the work on ``dueling bandits" \citep[starting from][]{Yue-dueling12,Yue-dueling-icml09}.
In this work, there is only one bandit algorithm which chooses two arms in each round, and the only observes which arm has ``won the duel".}
Dedicated monographs \citep{Bubeck-survey12,slivkins-MABbook,LS19bandit-book} cover the work on regret-minimizing formulations (which mainly comes from computer science). The classic book
\citep{Gittins-book11} focuses on the Markovian formulations (which predate regret-minimization and were mainly studied in operations research and statistics).
Various connections to economics and game theory are detailed in
books \citep{CesaBL-book,slivkins-MABbook} and surveys \citep{Bergemann-survey06,Horner-survey16}. Industrial applications are discussed in \citep{DS-arxiv}.

The three-way tradeoff between exploration, exploitation and incentives has been studied in several scenarios very different from ours:
incentivized exploration in recommendation systems
(starting from \citet{Kremer-JPE14,Che-13}, see \citet[Ch. 11.6]{slivkins-MABbook} for a literature review),
    %\citep[\eg][]{Che-13,Frazier-ec14,Kremer-JPE14,ICexploration-ec15,Bimpikis-exploration-ms17,Bahar-ec16,Jieming-unbiased18},
dynamic auctions
    (\eg see \citet{DynAuctions-survey10} for a survey),
    %\citep[\eg][]{AtheySegal-econometrica13,DynPivot-econometrica10,Kakade-pivot-or13},
pay-per-click ad auctions
    \citep[\eg][]{MechMAB-ec09,DevanurK09,Transform-ec10-jacm},
coordinating search and matching
    \citep{Bobby-Glen-ec16},
and human computation
    \citep[\eg][]{RepeatedPA-ec14,Ghosh-itcs13,Krause-www13}.
%A literature review of this work can be found in
%\citep[Ch. 11.6]{slivkins-MABbook}.

@@

The earlier work by \citet{bergemann1997market,bergemann2000experimentation} studies competition on both prices and quality, and allows for the data externality mentioned above. Specifically, the ``entrant" firm offers a new product of unknown, exogenously determined, quality, which is ``explored" if and only if the entrant attracts customers. However, because the entrant can only control the prices but not the product itself, the nature of exploration is fundamentally different than in our setting. First, the goal of exploration is to reveal information about a fixed product. Second, the entrant can only affect the \emph{quantity} of the said information, but not \emph{which} information is being revealed. Third, revealing more information is not necessarily better for the entrant (if the product's quality is actually low). Fourth, the social planner's exploration problem is very different from the firms', so that when one compares competition to the planner, one can only compare outcomes, but not the algorithms.\footnote{The planner faces a two-armed bandit problem, because the planner can directly choose a firm for each agent.}
% firm, because payments cancel out in social welfare).}

@@

%\gaedit{\citet{bergemann1997market,bergemann2000experimentation} studied a model where there is an incumbent with known product quality and an entrant with unknown, but exogenously determined, quality. Similar to \citet{keller2003price}, the firms compete on prices and they characterize the MPE of dynamic competition between the two firms. In their model, consumer purchases of the entrant's product result in an informational externality that allows both the firms and consumers to gather information on the entrant's product quality. However, we focus on a model where firms' product quality is endogenous and the focus of exploration in a context where the price mechanism is absent.}
In the line of work on \emph{strategic experimentation} (starting from \citet{Bolton-econometrica99,Keller-econometrica05}, see \citet{Horner-survey16} for a survey), agents explore and learn over time in a shared environment. Thus, we have exploration algorithms which interact with each other strategically, \eg each agent would prefer to free-ride on the exploration done by others. However, the agents do not compete with each other in any meaningful sense.

@@

\cite{schmalensee1982product, bagwell1990informational} investigate how buyer uncertainty about product quality can serve as a barrier to entry for late arrivers; we find a similar effect with ``reputation advantage". The role of data as a barrier to entry in online markets has been noted in \cite{de2020data}; we find a similar effect with ``data advantage". \citet{kerin1992first} overview the other channels through which first-mover advantage can lead to a competitive advantage.

@@

Classic ``market competitiveness" measures,such as the Lerner Index or the Herfindahl-Hirschman Index
\citep{tirole1988theory} are not applicable to our setting, as they rely on ex-post observable market attributes such as prices or market shares (which are, resp., absent and endogenous for us).

@@

\xhdr{Choice models.}
Stochastic choice models similar to ours have been widely used throughout economics. First, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. They play a similar role in our model, side-stepping the dominance of the greedy algorithm.
%as we move from \HardMax to \HardMaxRandom.
%\gadelete{Second, there a large literature on non-existence of equilibria due to small deviations, starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets. \footnote{\cite{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom in this context.} This is superficially similar to how small deviations towards the greedy algorithm rule out equilibria under \HardMaxRandom.}
Second, \SoftMaxRandom subsumes the logit choice rule, a standard behavioral model with strong empirical and microeconomic foundations
\citep[\eg][]{mosteller1951experimental, luce1959choice, mcfadden1974measurement, matvejka2015rational}.
Third, choice models similar to \SoftMaxRandom are used to explain  horizontal product differentiation \citep[\eg][]{Hotelling-29, Perloff-Salop-85}.
%\asdelete{While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}

>>>>

\xhdr{Exploration.} Multi-armed bandits (\emph{MAB}) is an elegant and tractable abstraction for tradeoff between \emph{exploration} and \emph{exploitation}: essentially, between acquisition and usage of information. MAB problems have been studied for many decades by researchers from computer science, operations research, statistics and economics, generating a vast and multi-threaded literature.  The most relevant thread concerns the basic model of regret-minimizating bandits with stochastic rewards and no auxiliary structure (which is the problem faced by each principal in our model), see Appendix~\ref{app:bg} for background. This basic model has been extended in many different directions, with a considerable amount of work on each: \eg payoffs with a specific structure (\eg combinatorial, linear, convex or Lipschitz), payoff distributions that change over time, and auxiliary payoff-relevant signals.
%\footnote{There is a superficial similarity, in name only, between this paper and the work on ``dueling bandits" \citep[starting from][]{Yue-dueling12,Yue-dueling-icml09}. In this work, there is only one bandit algorithm which chooses two arms in each round, and the only observes which arm has ``won the duel".}
Dedicated monographs \citep{Bubeck-survey12,slivkins-MABbook,LS19bandit-book} cover the work on regret-minimizing formulations (which mainly comes from computer science). The classic book
\citep{Gittins-book11} focuses on the Markovian formulations, which predate regret-minimization. Various connections to economics and game theory are detailed in
books \citep{CesaBL-book,slivkins-MABbook} and surveys \citep{Bergemann-survey06,Horner-survey16}. Industrial applications are discussed in \citep{DS-arxiv}.

The three-way tradeoff between exploration, exploitation and incentives has been studied in several scenarios very different from ours:
incentivized exploration in recommendation systems
(starting from \citet{Kremer-JPE14,Che-13}, see \citet[Ch. 11]{slivkins-MABbook}),
    %\citep[\eg][]{Che-13,Frazier-ec14,Kremer-JPE14,ICexploration-ec15,Bimpikis-exploration-ms17,Bahar-ec16,Jieming-unbiased18},
dynamic auctions
    (\eg see \citet{DynAuctions-survey10} for a survey),
    %\citep[\eg][]{AtheySegal-econometrica13,DynPivot-econometrica10,Kakade-pivot-or13},
pay-per-click ad auctions
    \citep[\eg][]{MechMAB-ec09,DevanurK09,Transform-ec10-jacm},
coordinating search and matching
    \citep{Bobby-Glen-ec16},
and human computation
    \citep[\eg][]{RepeatedPA-ec14,Ghosh-itcs13,Krause-www13}.
%A literature review of this work can be found in
%\citep[Ch. 11.6]{slivkins-MABbook}.

@@

The earlier work by \citet{bergemann1997market,bergemann2000experimentation} studies competition on both prices and quality, and allows for the data externality mentioned above. Specifically, the ``entrant" firm offers a new product of unknown, exogenously determined, quality, which is ``explored" if and only if the entrant attracts customers. However, because the entrant can only control the prices but not the product itself, the nature of exploration is fundamentally different than in our setting. First, the goal of exploration is to reveal information about a fixed product. Second, the entrant can only affect the \emph{quantity} of the said information, but not \emph{which} information is being revealed. Third, revealing more information is not necessarily better for the entrant (if the product's quality is actually low). Fourth, the social planner's exploration problem is very different from the firms', so that when one compares competition to the planner, one can only compare outcomes, but not the algorithms.\footnote{The planner faces a two-armed bandit problem, since it can directly choose a firm for each agent.}
% firm, because payments cancel out in social welfare).}

@@

%\gaedit{\citet{bergemann1997market,bergemann2000experimentation} studied a model where there is an incumbent with known product quality and an entrant with unknown, but exogenously determined, quality. Similar to \citet{keller2003price}, the firms compete on prices and they characterize the MPE of dynamic competition between the two firms. In their model, consumer purchases of the entrant's product result in an informational externality that allows both the firms and consumers to gather information on the entrant's product quality. However, we focus on a model where firms' product quality is endogenous and the focus of exploration in a context where the price mechanism is absent.}
In the line of work on \emph{strategic experimentation} (starting from \citet{Bolton-econometrica99,Keller-econometrica05}, see \citet{Horner-survey16} for a survey), agents explore and learn over time in a shared environment. Thus, we have exploration algorithms which interact with each other strategically, \eg each agent prefers to free-ride on someone else's exploration. However, the agents do not compete with each other in any meaningful sense.

@@

\cite{schmalensee1982product, bagwell1990informational} investigate how buyer uncertainty about product quality can serve as a barrier to entry for late arrivers; we find a similar effect with ``reputation advantage". \cite{de2020data} note the role of data as a barrier to entry in online markets; we find a similar effect with ``data advantage". \citet{kerin1992first} overview other channels through which first-mover advantage can \asedit{affect competition}.% lead to a competitive advantage.

@@

\asedit{While we use first-mover advantage and agents' decision rule,} classic ``market competitiveness" measures, such as the Lerner Index or the Herfindahl-Hirschman Index
\citep{tirole1988theory}, are not applicable to our setting, as they rely on ex-post observable market attributes such as prices or market shares (which are, resp., absent and endogenous for us).

@@

\xhdr{Choice models.}
Stochastic choice models similar to ours have been widely used throughout economics. First, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. They play a similar role in our model, side-stepping the dominance of the greedy algorithm.
%as we move from \HardMax to \HardMaxRandom.
%\gadelete{Second, there a large literature on non-existence of equilibria due to small deviations, starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets. \footnote{\cite{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom in this context.} This is superficially similar to how small deviations towards the greedy algorithm rule out equilibria under \HardMaxRandom.}
Second, \SoftMaxRandom subsumes the logit choice rule, a standard behavioral model with strong empirical and microeconomic foundations
\citep[\eg][]{mosteller1951experimental, luce1959choice, mcfadden1974measurement, matvejka2015rational}.
Choice models similar to \SoftMaxRandom are used to explain  horizontal product differentiation \citep[\eg][]{Hotelling-29, Perloff-Salop-85}.
%\asdelete{While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}

### 2021-01-18 journal paper/content/sec-intro.tex
Minor space-saving.

Interestingly, we find that data feedback loops can be bad for innovation (as in ``death spiral"), or good for innovation (as in ``amplification effect"), depending on the level of competition.

@@

One model we study is suitable for analysis and another for simulations, but not vice versa. A natural implementation of the \TheoryModel requires running time quadratic in the number of rounds,%
\footnote{\label{fn:Tsquared}\Eg this is because at each round $t$, one needs to recompute, and integrate over, a discrete distribution with $t$ possible values, namely the number of agents that have chosen principal $1$ so far.}
which precludes \gaedit{numerical simulations} at a sufficient scale. The \ExptsModel features an intricate feedback loop between algorithms' performance, their reputations and agents' choices, which simplifies the simulations but does not appear analytically tractable.

>>>>

Interestingly, 
\asdelete{we find that}
data feedback loops can be bad for innovation (as in ``death spiral"), or good for innovation (as in ``amplification effect"), depending on the level of competition.

%Interestingly, we find that data feedback loops can be bad for innovation (as in ``death spiral"), or good for innovation (as in ``amplification effect"), depending on the level of competition.

@@

One model we study is suitable for analysis and another for simulations, but not vice versa. A natural implementation of the \TheoryModel requires running time quadratic in the number of rounds,%
\footnote{\label{fn:Tsquared}\Eg this is because at each round $t$, one needs to recompute, and integrate over, a discrete distribution with $t$ possible values, namely the number of agents that have chosen principal $1$ so far.}
which precludes numerical simulations at a sufficient scale. The \ExptsModel features an intricate feedback loop between algorithms' performance, their reputations and agents' choices, which simplifies the simulations but does not appear analytically tractable.
